---
title: "Polychrome: Plots of Many Colors"
author: "Kevin R. Coombes"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polychrome}
  %\VignetteKeywords{OOMPA,Polychrome,Color Palettes,Palettes}
  %\VignetteDepends{Polychrome}
  %\VignettePackage{Polychrome}
  %\VignetteEngine{knitr::rmarkdown}
---

In this vignette, we describe how to use the palettes that ship with
the <tt>Polychrome</tt> package. We also describe the tools that are
provided to view palettes in different settings and different kinds
of plots.

## Getting Started

As usual, we start by loading the package:

```{r}
library(Polychrome)
```

Next, we can look at one of the color palettes it provides.

```{r}
mypal <- kelly.colors(22)
```

```{r fig.width=7, fig.height=5}
swatch(mypal)
```

## Resorted barplots

To get a better idea of whether colors are distinguishable, we can
sort them by hue:
```{r fig.width=7, fig.height=5}
swatchHue(mypal)
```

or by luminance:
```{r fig.width=7, fig.height=5}
swatchLuminance(mypal)
```

or in random order:
```{r fig.width=7, fig.height=5}
ranswatch(mypal)
```

## Scatter Plots
Some colors that work well in barplots may work less well in scatter
plots.  To test that possibility, we can generate some random points
and plot them. The next line of code generates 14 points from each
color in the "alphabet" palette.

```{r fig.width=7, fig.height=5}
pal2 <- alphabet.colors(26)
ranpoints(pal2, 14)
```

Our standard way to think about colors is in the L\*u\*v\* color space
model defined by the CIE.  The next plot shows how the points in the
palette are distributed in the u-v subspace.
```{r fig.width=7, fig.height=5}
uvscatter(pal2)
```

We can also show the associated luminance, L.
```{r fig.width=7, fig.height=5}
luminance(pal2)
```

## Plotting Curves
We can also test if a palette is useful for drawing a graph full of
different curves rarther than points.
```{r fig.width=7, fig.height=5}
rancurves(pal2)
```

## Color Similarity
We use L\*u\*v\* color space because perceptual distinguishability seems
to be closely related to Euclidean distance in this space.  To get
another view of which colors are most similar, we can perform
hierarchical clustering of colors using this representation.
```{r fig.width=7, fig.height=5}
plothc(pal2)
```

We can also perform principal components in L\*u\*v\* space to display
a projection that tries to preserve distances.
```{r fig.width=7, fig.height=5}
plotpc(pal2)
```

## Grayscale
One further consideration that should be kept in mind when creating
color palettes is that not every potential viewer has normal color
vision. One way to test if a color palette will remain useful in this
context is to convert it to gray scale. We do this by converting to
grayscale using the desaturate function from the colorspace package,
and then preparing a barplot by luminance. 

```{r fig.width=7, fig.height=5}
grayed <- colorspace::desaturate(pal2)
swatchLuminance(grayed)
```
