---
title: "Using Polychrome With ggplot"
author: "Kevin R. Coombes"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Polychrome With ggplot}
  %\VignetteKeywords{Polychrome,Color Palettes,Palettes,ggplot}
  %\VignetteDepends{Polychrome}
  %\VignettePackage{Polychrome}
  %\VignetteEngine{knitr::rmarkdown}
---

In this vignette, we describe how to use <tt>Polychrome</tt> palettes
with the package <tt>ggplot2</tt>. The vignette will only run code if
the <tt>ggplot2</tt> package is available.
```{r}
evalVignette <- requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = evalVignette)
```
## Getting Started
We want to build a custom palette of 40 colors for this application, with
each block of four consecutive colors being distinguishable. We start by
constructing a new palette in the usual way.

```{r newpal, fig.cap="A new palette of 40 colors.", fig.width = 8, fig.height=4}
library(Polychrome)
set.seed(935234)
P40 <- createPalette(40, c("#FF0000", "#00FF00", "#0000FF"), range = c(30, 80))
swatch(P40)
```
We achieve the goal of making the blocks of four colors being distinguishable
by first sorting by hue, and then rearranging them into four-blocks.
```{r reorder, fig.cap="A sorted palette of 40 colors.", fig.width = 8, fig.height=4}
P40 <- sortByHue(P40)
P40 <- as.vector(t(matrix(P40, ncol=4)))
swatch(P40)
```
Here is the **key point** of this entire vignette: By default,
<tt>Polychrome</tt> gives names to each of the colors in a palette.
But, in <tt>ggplot</tt>, named colors will only be applied if they
match the levels of an appropriate factor in the data. The simplest
solution is to remove the names:
```{r dename}
names(P40) <- NULL
```

## Simulating Complex Data
For illustration purposes, we simulate a data set with a moderately
complex structure. Specifically, we assume that we have

* Nine groups of samples,
* Four samples in each group,
* Three replicate experiments for each sample, and
* Measurements repeated over four days.

Here is the simulated design of the data set.
```{r simdata}
## Nine groups
NG <- 9
gp <- paste("G", 1:NG, sep = "")
length(gp)

## Four Subjects per group
## 36 Subjects = 9 groups * 4 subjects/group 
sid <- paste(rep(LETTERS[1:2], each=26), c(LETTERS, LETTERS), sep="")[1:(4*NG)]
length(sid)

## Three Reps per subject
## 108 Experiments
reps = factor(rep(c("R1", "R2", "R3"), times = length(sid)))
length(reps)

## Each experiment with measurements on four Days, so 432 data rows
daft <- data.frame(Day = rep(1:4, each=length(reps)),
                   Group = factor(rep(rep(gp, each=12), times = 4)),
                   Subject = factor(rep(rep(sid, each = 3), times=4)),
                   Rep = factor(rep(reps, times = 4)))
dim(daft)
summary(daft)
```

Now we add simulated "measurements" taken on each replicate of each subject
on each of four days.
```{r variable}
## Linear model with noise, ignoring group
beta <- runif(length(sid), 0.5, 2)
## "Measured" variable
attach(daft)
daft$variable <- rnorm(nrow(daft), 0, 0.2) + 1  + beta[as.numeric(Subject)]*Day
detach()
```

### Plotting the results.

```{r fig.cap="A faceted plot, colored by subject.", fig.width=8, fig.height=8}
library(ggplot2)
ggplot(daft, aes(x = Day, y = variable, colour = as.factor(Subject))) +
  geom_point(aes(shape = as.factor(Rep)), size = 3) +
  geom_line(aes(linetype = as.factor(Rep)), size = 0.8) +
  facet_wrap(. ~ Group, ncol = 3)+
  theme_bw() + theme(legend.position="none")+
  scale_color_manual(values = P40)
```

